### 2016-01-28 paper/final/source/Levin_JTB.tex
In process of drafting Response 3

\documentclass[preprint,10pt,authoryear]{elsarticle}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
%\usepackage{amssymb}

@@

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
\newcommand{\new}[1]{{\color{dkgreen}#1}}
\newcommand{\drew}[1]{{\color{dkgreen}#1}}
\newcommand{\fred}[1]{{\color{dkblue}#1}}
\newcommand{\steph}[1]{{\color{dkpurple}#1}}

@@

The model is parameterized with values taken from the literature when available.  Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and an earlier ODE model \citep{Mitchell2011}.   To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.   Of the cytokines tested, only the chemokines CXCL10 (IP-10) and CCL5 (RANTES) were expressed differently across the three influenza strains.   IP-10 and RANTES are both chemokines that normally attract T cells to the FOI \citep{Hoji2005, Groom2011a}.  These results suggest that the two chemokines may contribute to the immune system's varying ability to control different strains of influenza, and were thus included in our model.

The spatial model and the supporting experiments reveal several interesting and underappreciated phenomena.  The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection. Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration. The model suggests that the chemokine response of the host cells can hinder T cells’ ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain. A sensitivity analysis of the model parameters isolated the effects of each parameter in the model.  The analysis shows that most model parameters can vary across a broad range of values without strong effect on predicted outcomes.  Further analysis shows model predictions remain unchanged within broad categories of parameter type (e.g., all parameters related to chemokines, parameters related to viral kinetics, etc.).  For example, the chemokine-related parameters are stable in that they do not affect the model behavior within biologically plausible ranges. In contrast, parameters related to viral kinetics significantly alter model predictions.  Finally we examine the `window of control' which establishes an upper bound on the time it takes T cells to arrive at the FOI while still clearing the infection.  We find that this value is consistent with our sensitivity analysis and helps explain the challenge of controlling the highly virulent pH1N1 strain.

@@

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.  T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model. T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.

@@

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.   Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1 using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant.  Next, 1,000 bootstrapping runs were performed on each modeled strain to generate confidence intervals over $r$.  The results of these fits are shown in Table 1 (viral secretion rates are from previous fits in \cite{Mitchell2011}).  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).

@@

$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region. $T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum. The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$). The lung is modeled as a circular region with radius $R$. The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells. See Fig. 1 for more details about the model.  The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.  $a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.

@@

Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).  The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).  Consistent results between the two approaches suggests the less computationally intensive OFAT analysis is also accurate for aH5N1 and pH1N1.

@@

Although leukocytes exhibit directional behavior to chemokines \citep{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.   While T cell chemotaxis may not be proven, our previous work has shown severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.  Our sensitivity analysis shows that slowing T cell speed (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.   These results are consistent with the hypothesis that T cells are guided by chemokine in the lung epithelium.

@@

Previous ODE models have shown that viral dynamics are critical in determining the course of an infection \citep{Handel2008, Lee2009, Miao2010a, Mitchell2011, Saenz2010, Crauste2015, Price2015}.  The use of an ABM can complement spatially homogeneous differential equation models \citep{Beltman2007, Textor2014, Vroomans2012, Zheng2008}.  In this study, we tested how chemokine-directed T cell search contributes to infection clearance.  Explicitly modeling T cell chemotaxis revealed unique challenges to viral clearance brought about by infections that spread more rapidly than the chemokine gradient could decay and diffuse.  Furthermore, our model was also able to account for spatial constraints of viral diffusion and infection.  Figure \ref{fig:cycells} illustrates how new virus diffuses over cells that are no longer able to be infected.  ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.  Our model visualizes how the early infection has a higher proportion of virus overlapping healthy cells (Figure~\ref{fig:cycells}).

>>>>

\documentclass[preprint,10pt,numbers]{elsarticle}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

@@

%\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
%\newcommand{\new}[1]{{\color{dkgreen}#1}}
\newcommand{\removed}[1]{}
\newcommand{\new}[1]{#1}
\newcommand{\drew}[1]{{\color{dkgreen}#1}}
\newcommand{\fred}[1]{{\color{dkblue}#1}}
\newcommand{\steph}[1]{{\color{dkpurple}#1}}

@@

The model is parameterized with values taken from the literature when available.  Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and \removed{an earlier}\new{our previously published} ODE model \citep{Mitchell2011}.

To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.   Of the cytokines tested, only the chemokines CXCL10 (IP-10) and CCL5 (RANTES) were expressed differently across the three influenza strains.   IP-10 and RANTES are both chemokines that normally attract T cells to the FOI \citep{Hoji2005, Groom2011a}.  These results suggest that the two chemokines may contribute to the immune system's varying ability to control different strains of influenza, and were thus included in our model.

The \new{analyses of the} spatial model \removed{and the supporting experiments} reveal several interesting and underappreciated phenomena.  The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection. Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration. The model suggests that the chemokine response of the host cells can hinder T cells’ ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain. A sensitivity analysis of the model parameters isolated the effects of each parameter in the model.  The analysis shows that most model parameters can vary across a broad range of values without strong effect on predicted outcomes.  Further analysis shows model predictions remain unchanged within broad categories of parameter type (e.g., all parameters related to chemokines, parameters related to viral kinetics, etc.).  For example, the chemokine-related parameters are stable in that they do not affect the model behavior within biologically plausible ranges. In contrast, parameters related to viral kinetics significantly alter model predictions.  Finally we examine the `window of control' which establishes an upper bound on the time it takes T cells to arrive at the FOI while still clearing the infection.  We find that this value is consistent with our sensitivity analysis and helps explain the challenge of controlling the highly virulent pH1N1 strain.

@@

Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.  We extended \removed{a}\new{our} previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.  T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model. T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.

@@

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.   Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1 using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant.  Next, 1,000 bootstrapping runs \new{using resampled residuals} were performed on each modeled strain to generate confidence intervals over $r$.  The results of these fits are shown in Table 1 (viral secretion rates are from previous fits in \cite{Mitchell2011}).  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).

@@

$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region. $T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum. The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$). The lung is modeled as a circular region with radius $R$. The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells. See Fig. 1 for more details about the model.  The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.  $a$ and $b$ were fit \removed{using 12}\new{to 5} experimental runs of a spatial model implemented in CyCells \new{($R^2 = 0.997$, $P < 0.01$)}.

@@

Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain \new{as described in \cite{Marino2008}} using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).  The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).  Consistent results between the two approaches suggests the less computationally intensive OFAT analysis is also accurate for aH5N1 and pH1N1.

@@

Although leukocytes exhibit directional behavior to chemokines \citep{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.   While T cell chemotaxis may not be proven, our previous work \removed{has shown}\new{suggests} severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.  Our sensitivity analysis shows that slowing T cell speed (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.   These results are consistent with the hypothesis that T cells are guided by chemokine in the lung epithelium.

@@

\removed{Previous ODE models have shown that viral dynamics are critical in determining the course of an infection}\new{Previous ODE models have studied how viral dynamics affect the course of infection} \citep{Handel2008, Lee2009, Miao2010a, Mitchell2011, Saenz2010, Crauste2015, Price2015}.  The use of an ABM can complement spatially homogeneous differential equation models \citep{Beltman2007, Textor2014, Vroomans2012, Zheng2008}.  In this study, we tested how chemokine-directed T cell search contributes to infection clearance.  Explicitly modeling T cell chemotaxis revealed unique challenges to viral clearance brought about by infections that spread more rapidly than the chemokine gradient could decay and diffuse.  Furthermore, our model was also able to account for spatial constraints of viral diffusion and infection.  Figure \ref{fig:cycells} illustrates how new virus diffuses over cells that are no longer able to be infected.  ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.  Our model visualizes how the early infection has a higher proportion of virus overlapping healthy cells (Figure~\ref{fig:cycells}).
